---
title: "Project2_ver1"
author: "Bhavna Murthy"
date: "3/2/2021"
output: html_document
---
## Abstract

## Introduction
Is there a difference in case proportion among countries in the WHO region of Americas that implemented a mask mandate by April 1? If so did the countries with a mask policy in place by April 1 have lower case proportions?
## Background

## Decriptive Analysis
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(readxl)
library(dplyr)
library(ggplot2)
library(reshape2)
```


```{r}
covid <- read_csv("https://covid19.who.int/WHO-COVID-19-global-data.csv")

policy <- read_excel("Data/Gov_Responses2Covid19_last.xlsx")
```

```{r}
by_country <- policy %>%
  group_by(country)

policy2 <- by_country %>%
  subset(d=='31aug2020' | d=='01apr2020')

for (x in 1:nrow(policy2)) {
  if (policy2$d[x]=='31aug2020') {
    policy2$d[x]<-'2020-08-31' 
  }
  else {
    policy2$d[x]<-'2020-4-01'
  }
} 

covid2 <- covid %>%
  subset(Date_reported=='2020-08-31' | Date_reported=='2020-04-01') 

policy2$date<-as.Date(policy2$d, "%Y-%m-%d")

policy2<-policy2 %>%
  subset(select=-c(d))

df<-merge(x=policy2, y=covid2, by.x=c("geoid", "date"), by.y=c("Country_code", "Date_reported"))
df<-df %>%
  mutate(case_proportion=(Cumulative_cases/as.numeric(population_2019))) 

df

```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.


Adding case mortality column
```{r}
df$Case_mort = df$Cumulative_deaths/df$Cumulative_cases
```

```{r}
head(df)
```

#summary statistics
```{r}
# earliet date
min(covid$Date_reported)
```
```{r}
# Latest date
max(covid$Date_reported)
```

```{r}
# Halfway point
median(covid$Date_reported)
```

```{r}
summary(df$cases)
```

```{r}
summary(df$deaths)
```

```{r}
summary(df$Case_mort)
```


```{r}
df[which.max(df$Case_mort),]
```


```{r}
### CREATE VARIABLE FOR CASE PROPORTION GROWTH RATE BETWEEN APRIL AND AUGUST ###
### Growth rate = (current value - lagged value) / (lagged value)

#case.growth<-((df$case_proportion[which(df$date=='2020-08-31')]-df$case_proportion[which(df$date=='2020-04-01')])/df$case_proportion[which(df$date=='2020-04-01')]) 

#df$case.growth<-rep(case.growth, each=2)

case.growth<-((df$Cumulative_cases[which(df$date=='2020-08-31')]-df$Cumulative_cases[which(df$date=='2020-04-01')])/df$Cumulative_cases[which(df$date=='2020-04-01')]) 

df$case.growth<-rep(case.growth, each=2)
```

```{r}
### GENERATE BOXPLOT FOR COUNTRIES WHICH HAD MASK POLICY IN APRIL VERSUS NO MASK POLICY IN APRIL ###

df.maskapril<-subset(df, masks==1 & date=='2020-04-01')

df.augresult<-subset(df, df$country %in% df.maskapril$country & date=='2020-08-31')

mask.result<-summary(df.augresult$case_proportion)


df.nomaskapril<-subset(df, masks==0 & date=='2020-04-01')

df.augresult.nm<-subset(df, df$country %in% df.nomaskapril$country & date=='2020-08-31')

nomask.result<-summary(df.augresult.nm$case_proportion)

boxplot(df.augresult$case.growth, df.augresult.nm$case.growth,
        names=c('Mask policy in April', 'No mask policy in April'),
        ylim=c(0, 1000),
        main='Cumulative case growth rate from Apr to Aug based on Apr 1st mask policy')
```

```{r}
### GENERATE BOXPLOT FOR COUNTRIES WHICH HAD TESTING POLICY IN APRIL VERSUS ###
### NO TESTING POLICY IN APRIL ###

df.test.april<-subset(df, testing==1 & date=='2020-04-01')

df.augresult.test<-subset(df, df$country %in% df.test.april$country & date=='2020-08-31')

test.result<-summary(df.augresult.test$case.growth)


df.notest.april<-subset(df, testing==0 & date=='2020-04-01')

df.augresult.notest<-subset(df, df$country %in% df.notest.april$country & date=='2020-08-31')

nomass.result<-summary(df.augresult.notest$case.growth)

boxplot(df.augresult.test$case.growth, df.augresult.notest$case.growth,
        names=c('Testing Policy in April', 'No Testing Policy in April'),
        ylim=c(0, 1000),
        main='Cumulative case growth rate from Apr to Aug based on Apr 1st testing policy')
```

```{r}
### GENERATE BOXPLOT FOR COUNTRIES WHICH HAD LOCKDOWN POLICY IN APRIL VERSUS ###
### NO LOCKDOWN POLICY IN APRIL ###

df.ld.april<-subset(df, domestic==1 & date=='2020-04-01')

df.augresult.ld<-subset(df, df$country %in% df.ld.april$country & date=='2020-08-31')

ld.result<-summary(df.augresult.ld$case_proportion)


df.nold.april<-subset(df, domestic==0 & date=='2020-04-01')

df.augresult.nold<-subset(df, df$country %in% df.nold.april$country & date=='2020-08-31')

nold.result<-summary(df.augresult.nold$case_proportion)

boxplot(df.augresult.ld$case.growth, df.augresult.nold$case.growth,
        names=c('Lockdown Policy in April', 'No Lockdown Policy in April'),
        ylim=c(0,1000),
        main='Cumulative case growth rate from Apr to Aug based on Apr 1st lockdown policy')

```


```{r}
april_masking = subset(df$masks, df$date == '2020-04-01')
august_case_prop = subset(df$case_proportion, df$date =='2020-08-31' )
august_case_mort = subset(df$Case_mort, df$date==  '2020-08-31')
ggplot()+
  geom_point(mapping = aes(x= august_case_prop, y = august_case_mort, fill = april_masking)) +
  xlab("Case proportion in August") +
  ylab("Case mortality rate in August") 
```


